The illusory promise of the Aligned Rank Transform

### 2025-06-27 text/20-background.qmd
small correction

Even in this very small dataset ($n = 2$), the three ART rankings are distinct from one another and also differ significantly from the ranking produced by a simple rank transformation. Interestingly, identical responses (e.g., the two values of $0.3$) can be assigned very different ranks, while responses that differ substantially (e.g., $0.3$ and $3.0$) may receive the same rank. More surprisingly, larger values can receive lower ranks —-- revealing that ART is a non-monotonic transformation. For instance, for subject $s_1$ under condition $b_1$, the responses are $0.4$ and $0.8$ for $a_1$ and $a_2$, yet their respective $ART_A$ ranks are $7.5$ and $3.0$. In this case, ART significantly distorts the actual difference between the two groups, $a_1$ and $a_2$.

As we will show later, this behavior can lead to unstable results and misleading conclusions, and it represents a central flaw of the method. For example, running a repeated-measures ANOVA on these ART ranks yields a *p*-value of $.044$ for the effect of $A$. In contrast, using a repeated-measures ANOVA on the ranks of simple rank transformation produces a *p*-value of $.46$, while applying a log-transformation --- more appropriate for this type of data --- yields a *p*-value of $.85$.

>>>>

Even in this very small dataset ($n = 2$), the three ART rankings are distinct from one another and also differ significantly from the ranking produced by a simple rank transformation. Interestingly, identical responses (e.g., the two values of $0.3$) can be assigned very different ranks, while responses that differ substantially (e.g., $0.3$ and $3.0$) may receive the same rank. More surprisingly, larger values can receive lower ranks —- revealing that ART is a non-monotonic transformation. For instance, for subject $s_1$ under condition $b_1$, the responses are $0.4$ and $0.8$ for $a_1$ and $a_2$, yet their respective $ART_A$ ranks are $7.5$ and $3.0$. In this case, ART significantly distorts the actual difference between the two groups, $a_1$ and $a_2$.

As we will show later, this behavior can lead to unstable results and misleading conclusions, and it represents a central flaw of the method. For example, running a repeated-measures ANOVA on these ART ranks yields a *p*-value of $.044$ for the effect of $A$. In contrast, using a repeated-measures ANOVA on the ranks of simple rank transformation produces a *p*-value of $.46$, while applying a log-transformation -- more appropriate for this type of data -- yields a *p*-value of $.85$.
